Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-20 paper.tex
Updates to the mutations section narrative.

@@

\subsection*{Finite sites mutations}
\label{sec-mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

\subsection*{Selection}
Selection was added to the coalescent ... Lots of simulators exist:
SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs},
msms~\citep{ewing2010msms},
% What about Cosi, did it simulate selection?
Cosi2~\citep{shlyakhter2014cosi2},
and discoal~\citep{kern2016discoal}.

@@

We have implemented the basic infrastructure exists for the structured
coalescent, which makes adding support for, e.g.,
inversions straightforward~\citep{peischl2013sequential}

@@

\begin{figure}
\begin{center}
% \includegraphics[width=6cm]{msprime-argon-runtimes.png}
\end{center}
\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
simulations with ARGON. To ensure the comparison with ARGON is
as fair as possible we run it with a mutation rate of 0 and with
the minimum output options (so that it writes out only an empty VCF file)}
\end{figure}

@@

\section*{Acknowledgments}
JK is supported by the Robertson Foundation.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
[Others fill in their ACKs please]

>>>>

%%Affiliations:
% Franz Baumdicker:
% Cluster of Excellence "Controlling Microbes to Fight Infections", Mathematical and Computational Population Genetics, University of Tübingen, Germany

@@

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because we are usually simulating neutral evolution in coalescent simulations,
the problem of simulating generating variation can be decomposed into
two independent steps:
first simulating the genetic ancestry (the trees), and then simulating
variation by superimposing neutral mutation processes on those trees.
The classical \texttt{Seq-Gen}~\citep{rambaut1997seq} program allows a range
of different models of sequence evolution to be imposed on a given
tree in Newick format. Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but in several cases including a range of different
models of sequence evolution
built-in~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal}.
The ultimate output of the simulations of trees and mutations is
then some representation of the resulting nucleotide sequences,
and information about the simulated mutations is lost.

\jkcomment{stuff to cover somewhere in this section}
\begin{itemize}
\item Tskit's approach of storing the mutations is a very simple idea,
but has profound computational consequences. We can compute statistics
about the sequences \emph{without ever seeing those sequences}.
\item Mutation simulations is only included in msprime for historical reasons.
It would be better to see msprime's mutation simulation as a separate
package. The mutation and ancestry code in msprime have no priveledged
access to each other, and mutation simulation would be no less efficient
if done in a different package.
\item Really emphasise this point: you can use msprime's mutation simulation
on \empH{any} tskit tree sequence, produced by SLiM, fwdpp, etc or
even tsinfer. It's completely modular.
\end{itemize}

Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

\subsection*{Selection}
Another ellaboration of the standard neutral coalescent with recombination
is the addition of selective sweeps~\citep{kaplan1989hitchhiking,braverman1995hitchhiking,kim2002detecting}.
In the context of the coalescent this amounts to creating a structured population
during the sojourn of the beneficial mutation through the population (i.e., the sweep phase)
in which lineages may transit between favored and unfavored backgrounds through
recombination. This allows for many selective sweep scenarios, including recurrent, partial,
and soft selective sweeps to be simulated efficiently, however in general this
efficiency comes at the cost of flexibility in comparison to forwards in
time simulation.

Currently there are a small number of other simulators that will generate
coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}.
Each of these packages has their own benefits and possible drawbacks,
however none of them at present will output tree sequence formatted data.

@@

We have implemented selective sweeps in \msprime\ similarly to
the \texttt{discoal} implementation-- sweep trajectories are
generated from a jump process approximation to the conditional diffusion
of an allele bound for fixation~\citep{coop2004ancestral}.
This implementation allows us to further generalize sweep models to
changing population sizes as well as selection from standing variation,
both of which are features that are planned for the near future.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for, e.g.,
inversions is straightforward~\citep{peischl2013sequential}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/dtwf-perf}
\end{center}
\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
simulations with ARGON. To ensure the comparison with ARGON is
as fair as possible we run it with a mutation rate of 0 and with
the minimum output options (so that it writes out only an empty VCF file)}
\end{figure}

@@

\section*{Acknowledgments}
JK is supported by the Robertson Foundation.
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
ADK was supported by NIH awards R01GM117241 and R01HG010774.
Franz Baumdicker is funded by the Deutsche Forschungsgemeinschaft
EXC 2064/1 -- Project number 390727645, and EXC 2124 -- Project number 390838134.
